Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-26 Thesis/04-chap4.qmd
final

Regression analysis is a fundamental statistical technique widely used for modelling data from many fields. To diagnose the fit of a model it is recommended that the residuals are plotted. If the fit is good, any variation remaining should be noise, consistent with sampling from a distribution specified by the error model. Deviations that might be observed from a residual plot are non-normality, heteroscedasticity, and other associations with the fitted values. @sec-first-paper established that visual methods for assessing residuals are superior to conventional testing in that they are not so sensitive to small departures. @sec-second-paper provides a computer vision model to alleviate the human labor needed to visually assessing residual plots. The next step is to deliver the use of the computer vision model to potential users, so that it can be widely used to the benefit of the analytics community.

@@

To be able to predict VSS for a residual plot, both fitted values and residuals are needed to be extracted from the regression model object supplied by the user. In R, statistical models like `lm` (linear model) and `glm` (generalized linear model) typically support the use of generic functions such as `fitted()` and `resid()` to retrieve these values. The `get_fitted_and_resid()` method, called by the checker, relies on these generic functions by default. However, generic functions only work with classes that have appropriate method implementations. Some regression modelling packages may not fully adhere to the `stats` package guidelines for implementing these functions. In such cases, overriding the method becomes necessary.

>>>>

Regression analysis is a fundamental statistical technique widely used for modeling data from many fields. To diagnose the fit of a model it is recommended that the residuals are plotted. If the fit is good, any variation remaining should be noise, consistent with sampling from a distribution specified by the error model. Deviations that might be observed from a residual plot are non-normality, heteroscedasticity, and other associations with the fitted values. @sec-first-paper established that visual methods for assessing residuals are superior to conventional testing in that they are not so sensitive to small departures. @sec-second-paper provides a computer vision model to alleviate the human labor needed to visually assessing residual plots. The next step is to deliver the use of the computer vision model to potential users, so that it can be widely used to the benefit of the analytics community.

@@

To be able to predict VSS for a residual plot, both fitted values and residuals are needed to be extracted from the regression model object supplied by the user. In R, statistical models like `lm` (linear model) and `glm` (generalized linear model) typically support the use of generic functions such as `fitted()` and `resid()` to retrieve these values. The `get_fitted_and_resid()` method, called by the checker, relies on these generic functions by default. However, generic functions only work with classes that have appropriate method implementations. Some regression modeling packages may not fully adhere to the `stats` package guidelines for implementing these functions. In such cases, overriding the method becomes necessary.

### 2024-08-26 Thesis/05-chap5.qmd
final

Principles of transparency and reproducible research have guided the work with all source code available in public and private repositories at <https://github.com/TengMCing>.  The thesis is written using Quarto [@Allaire_Quarto_2024] and is available online at <https://patrick-li-thesis.netlify.app>.

>>>>

Principles of transparency and reproducible research have guided the work with all materials related to the thesis at <https://github.com/TengMCing/PhD>.  The thesis is written using Quarto [@Allaire_Quarto_2024] and is available online at <https://patrick-li-thesis.netlify.app>.

### 2024-08-26 Thesis/A-appA.qmd
final

The analysis is conducted by setting up several scenarios, where the $\alpha$ is under or overestimated by 12.5%, 25% and 50%. Using the adjusted $\hat{\alpha}$, we recalculate the $p$-value for every lineup and show the results in Figure \ref{fig:sensitivity}. It can be observed that there are some changes to $p$-values, especially when the $\hat{\alpha}$ is multiplied by 50%. However, Table \ref{tab:sensitivity} shows that adjusting $\hat{\alpha}$ will not result in a huge difference in rejection decisions. There are only a small percentage of cases where the rejection decision change. It is very unlikely the downstream findings will be affected because of the estimate of $\alpha$.

@@

```{r}
upper_lower_p %>%
  left_join(vi_survey %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  filter(x_dist == "uniform") %>%
  group_by(unique_lineup_id) %>%
  summarise(p_value = first(p_value))) %>%
  group_by(modified) %>%
  count(adjusted_reject = adjusted_p_value <= 0.05, reject = p_value <= 0.05) %>%
  complete(adjusted_reject, reject, fill = list(n = 0)) %>%
  ungroup() %>%
  filter(adjusted_reject != reject) %>%
  mutate(case = ifelse(!adjusted_reject, "reject to not reject", "not reject to reject")) %>%
  rename(multiplier = modified) %>%
  select(-adjusted_reject, -reject) %>%
  pivot_wider(names_from = case, values_from = n) %>%
  mutate(reject_not_reject_per = round(`reject to not reject`/279*100, 2)) %>%
  mutate(not_reject_reject_per = round(`not reject to reject`/279*100, 2)) %>%
  select(multiplier,
         `reject to not reject`,
         reject_not_reject_per,
         `not reject to reject`,
         not_reject_reject_per) %>%
  kableExtra::kable(format = "latex", 
                    booktabs = TRUE,
                    col.names = c("Multiplier",
                                  "Reject to not reject",
                                  "%",
                                  "Not reject to reject",
                                  "%"),
                    label = "sensitivity", 
                    caption = "Examining how decisions might change if $\\hat{\\alpha}$ was different. Percentage of lineups that where the $p$-value would switch to above or below 0.05, when $\\hat{\\alpha}$ is multiplied by a multiplier.",
                    linesep = c("", "", "\\addlinespace"))
```

@@

It would be interesting if a different conventional test significance level results in both the visual tests and conventional tests reject only the same residual plots, and fails to reject the same residual plots. This would be a state where both systems agree perfectly. Figure \ref{fig:rej-rates} examines this. Plot A shows the percentage of residual plots rejected by the visual test, given the conventional test rejected (solid lines) or failed to reject (dashed lines). The vertical grey line marks significance level 0.05. When the significance level gets smaller, it is possible to see that the visual tests reject (nearly) 100% of the time that the conventional test rejects. However, there is no agreement, because the visual tests also increasingly reject residual plots where the conventional test failed to reject. Plot B is comparable to an ROC curve, where the percentage visual test rejection conditional on conventional test decision is plotted: Reject conditional on reject is plotted against reject conditional on fail to reject, for different significance levels. The non-linearity pattern results are close to being ideal, that the percentage of reject relative to fail to reject increases very slowly as the reject relative to reject converges to 100. The heteroskedasticity pattern is more problematic, and shows that the cost of rejecting less with the conventional test is disagreement with the visual test.

>>>>

The analysis is conducted by setting up several scenarios, where the $\alpha$ is under or overestimated by 12.5%, 25% and 50%. Using the adjusted $\hat{\alpha}$, we recalculate the $p$-value for every lineup and show the results in @fig-sensitivity. It can be observed that there are some changes to $p$-values, especially when the $\hat{\alpha}$ is multiplied by 50%. However, @tbl-sensitivity shows that adjusting $\hat{\alpha}$ will not result in a huge difference in rejection decisions. There are only a small percentage of cases where the rejection decision change. It is very unlikely the downstream findings will be affected because of the estimate of $\alpha$.

@@

```{r}
#| tbl-cap: "Examining how decisions might change if $\\hat{\\alpha}$ was different. Percentage of lineups that where the $p$-value would switch to above or below 0.05, when $\\hat{\\alpha}$ is multiplied by a multiplier."
#| label: tbl-sensitivity
upper_lower_p %>%
  left_join(vi_survey %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  filter(x_dist == "uniform") %>%
  group_by(unique_lineup_id) %>%
  summarise(p_value = first(p_value))) %>%
  group_by(modified) %>%
  count(adjusted_reject = adjusted_p_value <= 0.05, reject = p_value <= 0.05) %>%
  complete(adjusted_reject, reject, fill = list(n = 0)) %>%
  ungroup() %>%
  filter(adjusted_reject != reject) %>%
  mutate(case = ifelse(!adjusted_reject, "reject to not reject", "not reject to reject")) %>%
  rename(multiplier = modified) %>%
  select(-adjusted_reject, -reject) %>%
  pivot_wider(names_from = case, values_from = n) %>%
  mutate(reject_not_reject_per = round(`reject to not reject`/279*100, 2)) %>%
  mutate(not_reject_reject_per = round(`not reject to reject`/279*100, 2)) %>%
  select(multiplier,
         `reject to not reject`,
         reject_not_reject_per,
         `not reject to reject`,
         not_reject_reject_per) %>%
  kableExtra::kable(format = "latex", 
                    booktabs = TRUE,
                    col.names = c("Multiplier",
                                  "Reject to not reject",
                                  "%",
                                  "Not reject to reject",
                                  "%"),
                    linesep = c("", "", "\\addlinespace"))
```

@@

It would be interesting if a different conventional test significance level results in both the visual tests and conventional tests reject only the same residual plots, and fails to reject the same residual plots. This would be a state where both systems agree perfectly. @fig-rej-rates examines this. Plot A shows the percentage of residual plots rejected by the visual test, given the conventional test rejected (solid lines) or failed to reject (dashed lines). The vertical grey line marks significance level 0.05. When the significance level gets smaller, it is possible to see that the visual tests reject (nearly) 100% of the time that the conventional test rejects. However, there is no agreement, because the visual tests also increasingly reject residual plots where the conventional test failed to reject. Plot B is comparable to an ROC curve, where the percentage visual test rejection conditional on conventional test decision is plotted: Reject conditional on reject is plotted against reject conditional on fail to reject, for different significance levels. The non-linearity pattern results are close to being ideal, that the percentage of reject relative to fail to reject increases very slowly as the reject relative to reject converges to 100. The heteroskedasticity pattern is more problematic, and shows that the cost of rejecting less with the conventional test is disagreement with the visual test.

### 2024-08-26 Thesis/abstract.qmd
final

The inclusion of co-authors reflects the fact that the work came from active collaboration between researchers and acknowledges input into team-based research. In the case of [Chapter 2](#sec-first-paper), my contribution to the work involved the following:

```{r}
#| message: false
#| echo: false
library(tidyverse)
library(knitr)
library(kableExtra)
dt <- tribble(
  ~`Chapter`, ~`Publication title`, ~Status, ~`Student contribution`, ~`Co-authors contribution`, ~`Coauthors are Monash students`,
  2, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol", "Published in the Journal of Computational and Graphical Statistics", "80%  Concept, Analysis, Software, Writing", "D. Cook 10%,  \n E. Tanaka 5%, \n S. VanderPlas 5%", "No",
  3, "Automated Assessment of Residual Plots with Computer Vision Models", "To be submitted", "80% Concept, Analysis, Software, Writing", "D. Cook 10%,  \n P. Menéndez 5%, \n U. Laa 2.5%,  \n N. Langrené 2.5%", "No",
  4, "Software for Automated Residual Plot Assessment: autovi and autovi.web", "To be submitted", "80% Concept, Analysis, Software, Writing", "D. Cook 10%,  \n E. Tanaka 5%, \n S. VanderPlas 2.5%, K. Ackermann 2.5%", "No"
)

@@

>>>>

The inclusion of co-authors reflects the fact that the work came from active collaboration between researchers and acknowledges input into team-based research. In the case of [Chapter 2](#sec-first-paper), [Chapter 3](#sec-second-paper) and [Chapter 4](#sec-third-paper), my contribution to the work involved the following:

```{r}
#| message: false
#| echo: false
library(tidyverse)
library(knitr)
library(kableExtra)
dt <- tribble(
  ~`Chapter`, ~`Publication title`, ~Status, ~`Student contribution`, ~`Co-authors contribution`, ~`Coauthors are Monash students`,
  2, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol", "Published in the Journal of Computational and Graphical Statistics", "80%  Concept, Analysis, Software, Writing", "D. Cook 10%,  \n E. Tanaka 5%, \n S. VanderPlas 5%", "No",
  3, "Automated Assessment of Residual Plots with Computer Vision Models", "To be submitted", "80% Concept, Analysis, Software, Writing", "D. Cook 10%,  \n E. Tanaka 5%, \n S. VanderPlas 2.5%, K. Ackermann 2.5%", "No",
  4, "Software for Automated Residual Plot Assessment: autovi and autovi.web", "To be submitted", "80% Concept, Analysis, Software, Writing", "D. Cook 10%,  \n E. Tanaka 5%, \n S. VanderPlas 2.5%, K. Ackermann 2.5%", "No"
)

@@

\clearpage
